\pagestyle{empty}

\boldsubsection{Reading Quiz Section \ref*{sec:quant}}

\begin{enumerate}
  \item Let $P(x)$ be the proposition $x^2-1=0$, with domain all real numbers. Which of the following statements are true?
  \begin{enumerate}
    \item \makebox[110pt][l]{$P(1)$ \hfill (b)} \  \makebox[110pt][l]{$P(-1)$ \hfill (c)} \  \makebox[110pt][l]{$P(3)$ \hfill (d)} \ $P(x)$
    \setcounter{enumii}{4}
    \item \makebox[110pt][l]{$\forall x, P(x)$ \hfill (e)} \ \makebox[110pt][l]{$\exists x, P(x)$ \hfill (f)} \  $\neg\bigl(\forall x, P(x)\bigr)$
  \end{enumerate}
  
  \item A value $x_0$ in the domain of $P$ for which $P(x_0)$ is \emph{false} is known as a(n) \underline{\phantom{counterexample}\qquad\qquad}
  \begin{enumerate}
    \item \makebox[180pt][l]{example \hfill (b)} \ counterexample
    \setcounter{enumii}{2}
    \item \makebox[180pt][l]{realization \hfill (d)} \ solution
  \end{enumerate}
  
  \item Which of the following are equivalent to the given expression? 
  \[
    \neg\bigl(\forall x, \exists y, P(x,y)\bigr)
  \]
  \begin{enumerate}
    \item \makebox[180pt][l]{$\exists x, \forall y, P(x,y)$ \hfill (b)} \ $\neg \bigl(\exists x, \forall y, P(x,y)\bigr)$
    \setcounter{enumii}{2}
    \item \makebox[180pt][l]{$\exists x, \forall y, \neg P(x,y)$\hfill (d)} \ $\forall x, \exists y, \neg P(x,y)$
  \end{enumerate}
  
  \item True or False: the order of quantifiers in an expression can always be switched without changing the meaning of the expression.
\end{enumerate}


\boldsubsection{Practice Problems Section \ref*{sec:quant}}

\begin{enumerate}
	\item Write each of the following using propositional functions and quantifiers. Make sure to define any propositional functions you are using.
  \begin{enumerate}
   	\item Every class has an instructor.
		\item For all real numbers $x$ and $y$, if $x$ and $y$ are positive, then there exists a positive integer $n$ such that $nx > y$.
		\item For each positive integer $n$, there exists a real number which is positive and is less than $\frac 1n$.
  \end{enumerate}
  
  \href{https://youtu.be/6XgjXmTefw8}{Video Solution}
  
  \item Negate each proposition (a), (b), (c) in the previous problem.
  
  \href{https://youtu.be/wXZ72O47aV8}{Video Solution}
  
  \item Which of the following propositions are true and which false? Justify your answers.
	\begin{enumerate}
	  \item $\forall x\in\R,\exists y\in\R,\ y^4=4x$
	  \item $\exists y\in\R,\forall x\in\R,\ y^4=4x$
	  \item $\forall y\in\R,\exists x\in\R,\ y^4=4x$
	  \item $\exists x\in\R,\forall y\in\R,\ y^4=4x$
	\end{enumerate}
	
	\href{https://youtu.be/lfI_1HFK7uc}{Video Solution}
\end{enumerate}