\pagestyle{empty}

\boldsubsection{Reading Quiz Section \ref*{sec:wellorder}}

\begin{enumerate}
	\item Which of the following statements are true? Select all that apply.
	\begin{enumerate}
    \item Every well-ordered set of real numbers has a minimum element.
    \item If a set of real numbers has a minimum element, then it is well-ordered.
    \item Any finite set of real numbers is well-ordered.
    \item Induction proofs must have a base case of 0 or 1.
	\end{enumerate}
	
	
	\item True or false: the set $A=\{\frac 1n:n\in\N\}$ is well-ordered.
	
	
		
	\item For each $n\in\N$, let $P(n)$ and $Q(n)$ be propositions.
	\begin{enumerate}
		\item Let $s$ be the smallest natural number such that $P(s)$ is false. What can you say about the elements of the set $A=\{n\in\N:n<s\}$ with respect to the property $P$?
		\item Let $a$ be minimal such that $P(a)\vee Q(a)$ is false. What can you say about the elements of the set $B=\{n\in\N:n<a\}$ with respect to the properties $P$ and $Q$?
		\item Let $u$ be minimal such that $P(u)\wedge Q(u)$ is false. What can you say about the elements of the set $C=\{n\in\N:n<u\}$ with respect to the properties $P$ and $Q$?
		\item Assume that $P(1)$ is true, but that ``$\forall n\in\N,\ P(n)$'' is false. Explain why there exists a natural number $k$ such that the implication $P(k)\Longrightarrow P(k+1)$ is \emph{false.}
	\end{enumerate}
	
	
	\item Here is an argument attempting to justify $\sum\limits_{i=1}^ni=\frac 12n(n+1)+7$. What is wrong with it?\smallbreak
	Assume that the statement is true for some fixed $n$. Then
	\[
		\sum_{i=1}^{n+1}i=\sum_{i=1}^ni+(n+1)=\frac 12n(n+1)+7+(n+1)=\frac 12(n+1)[(n+1)+1]+7
	\]
	hence the statement is true for $n+1$ and, by induction, for all $n\in\N$.

% 	\item The fact that $\N$ is well-ordered is considered a(n)
% 	\begin{enumerate}
%     \item theorem
%     \item opinion
%     \item axiom
%     \item proof
% 	\end{enumerate}

	%\item True or False: a finite set can be dense in $\R$.
\end{enumerate}


\boldsubsection{Practice Problems Section \ref*{sec:induction}}

\begin{enumerate}
	\item Prove that $n!>2^n$ for all $n \ge 4$.

	\href{https://youtu.be/xt3AgzI59LU}{Video Solution}

	%\item Fill in the details in the proof of Lemma \ref{lemm:rationaldensityprep}.

	%\href{https://youtu.be/n5v-6WAnj9M}{Video Solution}
\end{enumerate}

