\pagestyle{empty}

\boldsubsection{Reading Quiz Section \ref*{sec:cantor}}

\begin{enumerate}
	\item A set $A$ is \emph{countably infinite} (or \emph{denumerable}) if \rule{3cm}{0.15mm}. Select all that apply.
  \begin{enumerate}
    \item There exists a surjection from $\N$ onto $A$.
    \item There exists an injection from $\N$ into $A$.
    \item There exists a bijection between $A$ and $\Q$.
    \item There exists an injection from $A$ into $\N$ and no injection from $A$ into any finite set.
  \end{enumerate}
    
  \item True or False: if $A$ is a proper subset of $B$, then $A$ has strictly smaller cardinality than $B$.
\end{enumerate}


\boldsubsection{Practice Problems Section \ref*{sec:cantor}}

\begin{enumerate}
	\item Show that the set of \emph{perfect squares} $A=\{n^2:n\in\N\}$ is countably infinite.
  
	\item Find an injective function $f:\N\to (0,1)$.

	\item Let $a,b\in\R$ with $a<b$. Find a bijective function $g:(0,1)\to (a,b)$ and show that your function is bijective. Hence conclude that any two finite-length open intervals in $\R$ have the same cardinality.
	
	\item Let $B$ be a subset of $A$. Suppose $B$ is countably infinite and let $a\in A\setminus B$. Show that $B \cup\{a\}$ is countably infinite.\par
	(\emph{Hint: let $h:B\to\N$ be bijective: how might you construct $k:B\cup\{a\}\to\N$ bijective?})
	
	\item Let $A$ be a countably infinite set.
	\begin{enumerate}
	  \item Prove that the Cartesian product $A\times A$ is countably infinite.
	  \item By induction, prove that $\underbrace{A \times \cdots \times A}_{n \text{ times}}$ is countably infinite, for all $n\in\N$.
	\end{enumerate}
\end{enumerate}

