\pagestyle{empty}

\boldsubsection{Reading Quiz Section \ref*{sec:prop}}

\begin{enumerate}
  \item A \emph{tautology} is a proposition which \underline{\phantom{is always true}\qquad\qquad}
  \begin{enumerate}
      \item is false no matter what the truth value of its component propositions.
      \item is only true when all of its component propositions are true.
      \item is never false, no matter what the truth value of its component propositions.
      \item is built only using the connectives $\land, \lor$.
  \end{enumerate}
  
  \item A \emph{contradiction} is a proposition which \underline{\phantom{is always false}\qquad\qquad}
  \begin{enumerate}
      \item is false no matter what the truth value of its component propositions.
      \item is only true when all of its component propositions are false.
      \item is never false, no matter what the truth value of its component propositions.
      \item is built only using the connective $\neg$.
  \end{enumerate}
  
  \item The \emph{contrapositive} of the conditional $P\implies Q$ is the conditional \underline{\phantom{$\not Q\implies\not P$}\qquad\qquad}
  \begin{enumerate}
      \item $\neg P \implies Q$
      \item $\neg Q \implies \neg P$
      \item $Q \implies P$
      \item $P \implies \neg Q$
  \end{enumerate}

  \item True or False: The \emph{converse} of $P\implies Q$ is logically equivalent to $P\implies Q$.
  
  \item The \emph{negation} of the conditional ``if I study at least 25 hours per week, then I will be successful'' is the proposition \underline{\phantom{$\not Q\implies\not P$}\qquad\qquad}
  \begin{enumerate}
      \item ``I study at least 25 hours per week, but I am not successful.''
      \item ``Either I study less than 25 hours per week, or I am successful.''
      \item ``Either I study at least 25 hours per week, or I am not successful.''
      \item `If I am successful, then I will study at least 25 hours per week.''
  \end{enumerate}
  
  \item De Morgan's laws state that:
  \begin{gather*}
    \neg(P\vee Q) \text{ is logically equivalent to } (1) \underline{\phantom{(\neg P)\wedge (\neg Q)}\qquad}\\
    \neg(P\wedge Q) \text{ is logically equivalent to } (2) \underline{\phantom{(\neg P)\vee (\neg Q)\qquad}}
  \end{gather*}
  \begin{enumerate}
      \item (1) $\neg (P \implies Q)$, \quad (2) $(\neg P \lor \neg Q)$
      \item (1) $(\neg P \land \neg Q)$, \quad (2) $(P \lor Q)$
      \item (1) $(\neg P \land Q)$, \quad (2) $(P \lor \neg Q)$
      \item (1) $(\neg P \land \neg Q)$, \quad (2) $(\neg P \lor \neg Q)$
  \end{enumerate}
\end{enumerate}


\boldsubsection{Practice Problems Section \ref*{sec:prop}}

\begin{enumerate}
  \item Suppose that ``If Colin was early, then no-one was playing pool'' is a true statement.
		\begin{enumerate}
	  	\item What is its contrapositive of this statement? Is it true?
	  	\item What is the converse? Is it true?
	  	\item What can we conclude (if anything?) if we discover each of the following? \emph{Treat the two scenarios separately.}
			\begin{enumerate}
	  	  \item[(i)] Someone was playing pool.
	      \item[(ii)] Colin was late.
			\end{enumerate}
		\end{enumerate}
		% (From previous exercises)
		
		\href{https://youtu.be/BI_M1-OoMac}{Video Solution}
  
  \item Prove that $P \lor \neg Q$ is logically equivalent to $\neg P \implies (\neg P \land \neg Q)$.
  
  \href{https://youtu.be/NrbdGXAP6Ug}{Video Solution}
  
  \item Define the connective $\uparrow$ (called the \emph{Sheffer stroke}, or \emph{NAND}) by the following truth table:
  \[
  \begin{array}{cc|c}
P & Q & P \uparrow Q\\\hline
T & T & F\\
T & F & T\\
F & T & T\\
F & F & T
\end{array}
  \]
  \begin{enumerate}
      \item Prove $P \uparrow Q$ is logically equivalent to $\neg (P \land Q)$. 
      \item Find an expression built using only $P$ and the connective $\uparrow$ which is logically equivalent to $\neg P$.
      \item Find an expression built using only $P$, $Q$, and the connective $\uparrow$ which is logically equivalent to $P \land Q$.
  \end{enumerate}
  
  \href{https://youtu.be/bxYgvuXtcTM}{Video Solution}
  
\end{enumerate}